\documentclass[]{article}
\input{packages}
\input{definitions}
\begin{document}

\title{wxMaxima Files}
\author{Mauro Mongiardo}
\date{Today}
\maketitle


The files which start with 1 are for rectangular coordinates, 2 corresponds to cylindrical coordinates, 3 corresponds to spherical coordinates.


\section{Rectangular coordinates}


The most complete file is:
\begin{verbatim}
File: 1Pauli_nabla3rect.wxm
\end{verbatim}
Nablarect is the block that performs the nabla operator in rectangular coordinates.
Gradesep performs the grade extraction. 
It is suggested to start from here.

The file
\begin{verbatim}
File: 1Example3.10rect.wxm
\end{verbatim}
perform the gradient evaluation of a given function.

The file
\begin{verbatim}
File: 1Example3.11rect.wxm
\end{verbatim}
perform the div and curl evaluation of a given vector.
Conventional analysis is also used. 

The file
\begin{verbatim}
File: 1Gradient_rectangular.wxm
\end{verbatim}
perform the gradient evaluation of a givenfunction with Pauli matrices.
 
 
The file
\begin{verbatim}
File: 1nabla_rectangular.wxm
\end{verbatim}
perform the div and curl evaluation of a given vector with Pauli matrices.
 



\section{Cylindrical coordinates}

\subsection{Cylindrical Potential}
The most complete file is the following:
\begin{verbatim}
File: Nabla_cyl_4_blocks.wxm
\end{verbatim}

This file contains a block performing the nabla operator in cylindrical coordinates.
It also contains the Pauli matrices for rectangular and cylindrical coordinates.
It also contains a block Gradecyl which perform the grade operation in cylindrical coordinates.

At the end there are two examples. 
The first example deals with a charged line (Electrostatic field)
The second example deals with an infinite filament with current.

\subsection{Conventional and GA nabla for cylindrical coordinates}
\begin{verbatim}
File: 2Nabla_cyl.wxm
\end{verbatim}

This files shows how to the div , grad curl operation in conventional analysis and with Pauli matrices.



\subsection{example 3.10, exercise 3.10, exercise 3.13}
\begin{verbatim}
File: Esempio_3.10.wxm
\end{verbatim}

This file contains a block performing the nabla operator in cylindrical coordinates.
It also contains the Pauli matrices for rectangular and cylindrical coordinates.

It solves the example 3.10 of Ulaby which for a scalar function
\be
V = {V}_{0}\,{e}^{-2\,\rho}\,\mathrm{sin}\left( 3\,\phi\right) 
\ee
ask to evaluate the gradient.
Then performs the divergence and the curl in cylindrical coordinates.


\section{Spherical coordinates: Dipole}

The most complete file is 
\begin{verbatim}
File: 3Nabla_sph_2_blocks.wxm
\end{verbatim}
which contains a block for the nabla evaluation in spherical coordinates and a block for the grade retrieval.


A first example of application is given in 
\begin{verbatim}
File: 3Dipole_sph_static1.wxm
\end{verbatim}



The file
\begin{verbatim}
File: 3Nabla_sph_exc_2.wxm
\end{verbatim}
is an example for a particular vector.



\begin{verbatim}
File: Dipole_sph_1.wxm
\end{verbatim}

This file contains a block performing the nabla operator in spherical coordinates.
It also contains the Pauli matrices for rectangular and spherical coordinates.

We assume a potential of the type
\be
\frac{{e}^{-j\,k\,r}}{r} \Bz_0
\ee
which is the same of a short dipole.
Perform the nabla operator on this potential.

This file also contains a block Gradesph that performs the grade extraction.

The Electric field is evaluated from the potential as
\bea
\Phi & = & - \frac{v}{jk} \nabla \cdot \BA\nonumber \\
\BE & = & -j \, k \,  v \,  \BA - \nabla\Phi \nonumber
\eea


\end{document}